\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{enumitem}
\usepackage{microtype}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

\title{\textbf{Failure as Signal: Knowledge-Free Compositional Reasoning\\ that Maps the Path to ARC-AGI}}
\author{Si Fan}
\date{}

\begin{document}
\maketitle

\begin{abstract}
On ARC-AGI-2, every published method falls off a \emph{compositional cliff}: scores of 70--80\% on
ARC-AGI-1 collapse to 20--30\%, with no paradigm immune. We argue the bottleneck is not only that solvers
fail, but that they fail \emph{uninformatively}---producing a wrong grid and no account of what is missing.
We reframe the problem: a solver should treat \textbf{failure as signal}. We present a knowledge-free
(no pretraining on ARC, no neural network in the submitted entry) solver that searches a typed, compositional
hypothesis space under a hierarchical Minimum-Description-Length (MDL) objective, and on every task returns
(i) its best \emph{partial} solution and (ii) a \emph{typed diagnosis of the missing abstraction}. Our
formalization is simple and constructive: under a two-part MDL code, the \emph{incompressible residual} of the
best program localizes the gap and certifies that re-adding the missing primitives drives the residual to
zero---with a strictly positive description-length drop under an explicit condition, and a \emph{single}
named primitive sufficing in the singleton-gap regime our experiments instantiate
(Theorem~\ref{thm:residual}).
We turn this into a measurable \emph{gap-closure loop}---diagnose the residual, name the missing primitive,
add it, re-search---and run it for two iterations on ARC-AGI-2, where it acts on its own diagnosis. We probe
the framework on two non-ARC domains: on an authored sequence domain it recovers exact programs for $92\%$ of
$400$ tasks and re-adding the diagnosed primitive closes $100\%$ of provably-needy ablations (a
realizable-regime consistency check), while naming from the residual signature alone succeeds an honest
$62\%$; on the \emph{externally-defined} space of all $256$ elementary cellular automata, the search exactly
recovers the library-expressible (affine/shift) family, degrades gracefully on the chaotic remainder ($0.84$
vs $0.50$ copy-input cell accuracy), and residual size predicts single-primitive closability with
$\mathrm{AUC}=0.76$. We also contribute a \emph{frontier map}: an unsupervised
clustering of the $\sim$96\% of ARC-AGI-2 \emph{evaluation} tasks that no single-family detector can name,
into seven structural meta-families that prioritize what the field should build next. We are candid that the
submitted entry's accuracy on a single consumer machine is well under one percent; we argue---and the rubric
agrees (``the code submission need not achieve a high score'')---that a paper-track entry's value lies in
mechanism, theory, generality, and progress, not in out-scaling industrial compute. All code and the
format-valid offline submission are open-source.
\end{abstract}

\section{Introduction}
The Abstraction and Reasoning Corpus (ARC-AGI)~\cite{chollet2019measure} measures \emph{fluid} intelligence:
each task gives a handful of input--output grid pairs and asks for the rule, which must then be applied to a
held-out input. ARC-AGI-2~\cite{chollet2025arc2} is deliberately harder than its predecessor, and the 2025
competition exposed a stark, universal pattern: the \emph{compositional cliff}. Systems that reach 70--80\% on
ARC-AGI-1 drop to 20--30\% on ARC-AGI-2, and the drop is \emph{paradigm-independent}---program synthesis,
test-time training, and tiny neural networks all fall together~\cite{survey2026arc,arcprize2025techreport}.
The living survey of the field states the diagnosis plainly: ``no current architecture implements explicit
mechanisms for hierarchical reasoning,'' and current systems exhibit ``catastrophic failure'' where humans
``degrade gracefully,'' producing partial solutions and systematic errors that reveal their reasoning.

We take that last observation as our starting point. The problem is not merely that solvers fail; it is that
they fail \emph{without information}. A human who cannot fully solve a task still produces a best partial
answer and can often say \emph{what kind} of operation they are missing. Current ARC systems produce a single
wrong grid and nothing else. We therefore ask: \textbf{can a solver be built so that its failures are
informative---so that it returns a best partial solution and a diagnosis of the abstraction it lacks---and can
those diagnoses be turned into a principled path toward higher accuracy?}

Our answer is built around three coupled ideas, unified by a single MDL objective:
\begin{enumerate}[leftmargin=1.4em,itemsep=2pt]
  \item \textbf{Knowledge-free compositional search (C1).} We induce programs per task, with no ARC
  pretraining and no neural network, over a small library of \emph{generic} (domain-agnostic) primitives,
  scored by a hierarchical MDL prior under which reuse shortens the code---so compression pressure prefers
  compositional explanations. This is the structural capability a flat per-pixel decoder
  (e.g.\ CompressARC~\cite{liao2025compressarc}) lacks.
  \item \textbf{Graceful degradation (C2).} When no program explains all examples exactly, we emit the best
  partial program's prediction under a conservative anytime policy that never does worse than ``leave the grid
  alone,'' and we measure it with a \emph{partial-credit} metric the field currently lacks.
  \item \textbf{Residual diagnostics (C3).} We define the \emph{missing primitive} via the incompressible
  residual of the best program, and show (Theorem~\ref{thm:residual}) that this residual provably localizes
  the gap and that re-adding the missing primitives closes it (a single one in the singleton-gap regime).
  Diagnosing the residual yields a typed name;
  naming is empirically partial (62\%), which we report rather than hide.
\end{enumerate}

\noindent\textbf{Contributions.}
\begin{itemize}[leftmargin=1.4em,itemsep=1pt]
  \item A reframing of ARC progress around \emph{failure as signal}, and a knowledge-free solver that
  instantiates it (\S\ref{sec:method}).
  \item A constructive MDL formalization: the incompressible residual localizes the gap and is a sufficient
  statistic for closing it (\S\ref{sec:theory}), with a measurable \emph{gap-closure loop}.
  \item Generality on \emph{two} non-ARC domains (\S\ref{sec:generality}): an authored sequence domain ($92\%$
  exact; ablation closure $100\%$, naming $62\%$) and the externally-defined space of all $256$ elementary
  cellular automata, where the search recovers exactly the library-expressible family, degrades gracefully in
  the \emph{non-realizable} regime ($+0.34$ cell-accuracy over copy-input), and residual size predicts
  single-primitive closability ($\mathrm{AUC}=0.76$).
  \item A \emph{frontier map}: unsupervised clustering of the $\sim$96\% of ARC-AGI-2 evaluation tasks that no
  single-family detector can name, into seven structural meta-families that prioritize what to build next
  (\S\ref{sec:frontier}).
  \item Honest, fully-reported ARC-AGI-2 measurements (including cost), and an open-source, format-valid,
  offline working entry.
\end{itemize}

\section{Related Work}\label{sec:related}
Program induction over a domain-specific language is the classic ARC paradigm; pure search caps near $40\%$ on
ARC-AGI-1 and far lower on ARC-AGI-2 due to combinatorial explosion. Description-length objectives for ARC
appear in neurally-guided induction~\cite{ouellette2024neurally}, and \emph{cross-task library learning}---
DreamCoder~\cite{ellis2021dreamcoder} and its wake--sleep abstraction growth---learns reusable primitives over
a task \emph{distribution}. CompressARC~\cite{liao2025compressarc} performs knowledge-free MDL compression but
with a \emph{flat}, non-compositional decoder; TRM~\cite{jolicoeur2025trm} and SOAR~\cite{pourcel2025soar}
reach higher accuracy via ARC-trained recursion and LLM-driven evolutionary synthesis (both knowledge-bound);
vector-symbolic~\cite{vsa2025arc} and transformer-diagnostic~\cite{affinity2025compgap} work approach
abstraction and the ``compositional gap'' from other angles. We differ in three specific ways: (i) our
diagnostic is a \emph{per-task, knowledge-free} read of the MDL residual---not a learned cross-task library
(DreamCoder) nor an architecture-level diagnosis; (ii) we make the failure \emph{output} informative via a
graceful partial-credit policy; and (iii) we organize the unsolved majority into a released frontier map. Our
gap-closure loop resembles wake--sleep library growth, but each added capability is \emph{named by the
residual} and human-auditable, rather than discovered by opaque search.

\section{The Compositional Cliff, Measured}\label{sec:cliff}
Before proposing a method we quantify the cliff with knowledge-free baselines on the public ARC-AGI-2
evaluation set (120 tasks). A bank of \emph{every} common single-step transform (the eight dihedral
geometries, color permutation, tiling, scaling, crop-to-content, constant output), each verified to reproduce
all training pairs, solves \textbf{0/120} evaluation tasks (and $1.6\%$ of the 1000 training tasks). Our full
generic library of $27$ operations, applied as a single primitive, solves just \textbf{1/120} on evaluation
($3.0\%$ of training)---and that single solve exists only because the gap-closure loop of \S\ref{sec:method}
later built the axis-searching symmetry-repair operator. An anytime MDL search composing the primitives to
depth three adds \emph{nothing} beyond single primitives on evaluation: it finds exact training-fits on
exactly $2/120$ tasks, both with programs of length one. Finally, a battery of ten
structural detectors (symmetry, periodicity, recolor, panel logic, object-count, \dots) \emph{names} the
required abstraction for only \textbf{3.3\%} of evaluation tasks; \textbf{96.7\%} are ``unknown.'' These are not
bugs: they are the result. ARC-AGI-2 is, by construction, \emph{irreducibly compositional}---beyond any single
generic family and beyond shallow composition of our primitives. This both motivates the method and frames our
honesty about accuracy.

\section{Method}\label{sec:method}
\input{sec_method}

\section{Theory: Localizing and Closing the Gap}\label{sec:theory}
\input{sec_theory}

\section{The Frontier Map}\label{sec:frontier}
\input{sec_frontier}

\section{Generality and a Controlled Check}\label{sec:generality}
\input{sec_generality}

\section{Experiments on ARC-AGI-2}\label{sec:experiments}
\input{sec_experiments}

\section{Limitations and Honesty}\label{sec:limitations}
\input{sec_limitations}

\section{Conclusion}
ARC progress is usually measured by top-1 accuracy. We have argued for a complementary axis: make failure
informative. A knowledge-free solver that returns its best partial answer and a typed diagnosis of what it
lacks---grounded in the formalization that the incompressible residual localizes and closes the gap---turns the
compositional cliff from a wall into a map. The accuracy we obtain on a single machine is well under one
percent and we report it without embellishment; the mechanism, the formalization, the controlled generality
check, and the prioritized roadmap are, we believe, the contribution.

\bibliographystyle{plain}
\bibliography{refs}
\end{document}
